---
title: \vspace{-1cm} Analysis of Car Accidents in Barcelona
author: "Ivan Iliash, Johann Promeuschel, Manuel Coutinho, Onat Kaya"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Data Analysis and Visualization in R (IN2339) - Case Study
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

```{r libraries, include=FALSE}
library(ggplot2)
library(data.table)
library(dplyr)
library(tidyr)
library(ggrepel)
library(ppcor)
library(patchwork)
library(forcats)
library(viridis)
library(pROC)
```

```{r dts}
DATA_FOLDER <- 'data'

load_dt <- function(folder, file){
  dt <- fread(file.path(folder, file)) # Read document
  setnames(dt, colnames(dt), gsub(' ', '.', colnames(dt))) # Replaces spaces with . in column names
  return(dt)
}

acc_dt <- load_dt(DATA_FOLDER, 'accidents_2017_fixed.csv')
pop_dt <- load_dt(DATA_FOLDER, 'population.csv')
transport_dt <- load_dt(DATA_FOLDER, 'transports.csv')
bus_dt <- load_dt(DATA_FOLDER, "bus_stops.csv")
```

# Motivation and Goals

In 2017, with a population of about 1.621 million, the capital of Catalonia had an accident rate way above the national average (582 vs 219^[From Dirección General de Tráfico, Accessed 16 January 2022 <https://revista.dgt.es/es/noticias/nacional/2018/07JULIO/0712-Siniestralidad-2017.shtml>] accidents with victims per 100.000 inhabitants).
In an attempt of mitigating the high number of accidents that occurred in the city of Barcelona, we show a collection of different findings that we made during the analysis and try to interpret some of the results related with the city's population, amount of public transport in certain areas, as well as time and location of accidents.



```{r goals, results="hide"}
# National Accident with victims rate:
total_accidents_with_victims <- 102233 #from source [1]
total_population <- 46658447 # from Wikipedia
total_acc_rate_per100 <- total_accidents_with_victims / total_population * 1e5

# Barcelona Accidents with victims rate:
barc_accidents_with_victims <- nrow(acc_dt[Victims > 0])
barc_population <- sum(pop_dt[Year == 2017, Number])
barc_acc_rate_per100 <- barc_accidents_with_victims / barc_population * 1e5

sprintf("Total number of accidents with victims per 100k: %d", round(total_acc_rate_per100))
sprintf("Barcelona number of accidents with victims per 100k: %d", round(barc_acc_rate_per100))
```

# Data Preparation
During the usual data preparation needed to plot the data, we noticed that some district and neighborhood were marked as "Unkown", so we filled out these gaps in the 'accidents_2017_fixed.csv' dataset by adding the correct values. This was done with a Python script using the Longitude and Latitude coordinates of the accident and requesting the information from the OpenStreetMap API^[<https://wiki.openstreetmap.org/wiki/API>].

```{r data_prep, results='hide'}
# Proof of dataset cleaning
sprintf("Number of Unkown Districts before cleaning vs after: %d - %d", 
        nrow(load_dt(DATA_FOLDER, 'accidents_2017.csv')[District.Name == 'Unknown']), 
        nrow(acc_dt[District.Name == 'Unknown']))

# 2017 Population x Accident Aggregated by District or Neighborhood
pop2017_dt <- pop_dt[Year == 2017]

pop_by_district_dt <- pop2017_dt[, .(N.Pop = sum(Number)), by=District.Name]
pop_by_neighborhood_dt <- pop2017_dt[, .(N.Pop = sum(Number)), by=Neighborhood.Name]
pop_by_neighborhood_dt[Neighborhood.Name == "el Poble Sec", Neighborhood.Name := "el Poble-sec"]

# Accidents by Neighborhood and by District
acc_by_neighboorhood_dt <- acc_dt[, .(N.Accidents = .N), by=Neighborhood.Name]
acc_by_district_dt <- acc_dt[, .(N.Accidents = .N), by=District.Name]

# Bus Stop by Neighborhood
stops_by_neighborhood_dt <- bus_dt[, .(N.Bus.Stops = .N), by=Neighborhood.Name]

# Transports by Neighborhood and by District
transport_by_neighboorhood_dt <- transport_dt[, .(N.Transports = .N), by=Neighborhood.Name]
transport_by_district_dt <- transport_dt[, .(N.Transports = .N), by=District.Name]


# Full Neighborhood and District Dataset
full_neighborhood_dt <- Reduce(function(x, y){merge(x, y, all.x = TRUE)}, list(pop_by_neighborhood_dt, acc_by_neighboorhood_dt, transport_by_neighboorhood_dt, stops_by_neighborhood_dt))
full_neighborhood_dt[is.na(full_neighborhood_dt)] <- 0

full_district_dt <- Reduce(function(x, y){merge(x, y, all.x = TRUE)}, list(pop_by_district_dt, acc_by_district_dt, transport_by_district_dt))

```


# Data Analysis
Since Barcelona is the second most populated city in Spain^[From Wikipedia, Accessed 17 January 2022 <https://en.wikipedia.org/wiki/List_of_European_cities_by_population_within_city_limits>], what first comes to mind is that the  number of accidents is related to the number of people living in a district or neighborhood, as that inevitably leads to more traffic.
We start by simply showing the number of accidents by district:

```{r acc_distr, fig.height = 5, out.height="30%", fig.align = "center", fig.width=7}
# TODO: don't know if geom_text is needed, simplify graphic? - clean lines, background, ...
acc_dt %>% ggplot(aes(x=fct_infreq(District.Name))) + 
  geom_bar() +
  geom_text(stat='count', aes(label=..count..), vjust=-0.3) +
  labs(x = 'District', y = 'Number of Accidents') +
  theme(axis.text.x = element_text(angle = 30, vjust = 0.95, hjust=0.65)) 

```

One can immediately notice a clear outlier: the Eixample district. This is, in fact, the most populated, and most visited district in Barcelona, which could explain the exceptionally high number of accidents.
We proceed by testing this proposed correlation, comparing different Neighborhoods and Districts:

```{r corr_pop_x_acc, fig.width=13, message = FALSE}

p1 <- full_neighborhood_dt %>% 
  ggplot(aes(x=N.Pop,y=N.Accidents)) + geom_point(alpha = 0.5) + stat_smooth(method="lm") + 
  geom_text_repel(aes(label=Neighborhood.Name)) +
  labs(x = '', y = 'Number of Accidents', title = "Number of Accidents by Neighborhood Population") +
  theme(plot.title = element_text(size=14, margin=margin(b=14))) 

p2 <- full_district_dt %>% 
  ggplot(aes(x=N.Pop,y=N.Accidents)) + geom_point(alpha = 0.5) + stat_smooth(method="lm") + 
  geom_text_repel(aes(label=District.Name)) +
  labs(x='', y='', title = "Number of Accidents by District Population") +
  theme(plot.title = element_text(size=14, margin=margin(b=14)))

p1 + p2
grid::grid.draw(grid::textGrob("Population", y=0.05, x=0.52))

```

Although both these correlations are statistically significant (p-value lower than .05 in both cases of the Spearman's rank correlation test), we can again notice Eixample and some of its neighborhoods (like "la Dreta de l'Eixample" and "l'Antiga Esquerra de l'Eixample") as clear outliers. This could be explained by the centrality of this district, which further heightens the traffic. 

```{r corr_pop_x_acc_tests, results="hide"}

print("Population in neighboorhood vs N Accident")
cor.test(full_neighborhood_dt$N.Pop, full_neighborhood_dt$N.Accidents, method="spearman", exact=FALSE)

print("Population in district vs N Accident")
cor.test(full_district_dt$N.Pop, full_district_dt$N.Accidents, method="spearman")
```

As we don't have any direct metric for the traffic, we attempt use the number of transportation stations in a certain area as a proxy for its afflux of people.

```{r corr_trans_x_acc, fig.width=13, message = FALSE}

p1 <- full_neighborhood_dt %>% 
  ggplot(aes(x=N.Transports,y=N.Accidents)) + geom_point(alpha = 0.5) + stat_smooth(method="lm") + 
  geom_text_repel(aes(label=Neighborhood.Name)) +
  labs(x = '', y = 'Number of Accidents', title = "Number of Accidents against Transports in the Neighborhood") +
  theme(plot.title = element_text(size=14, margin=margin(b=14))) 

p2 <- full_district_dt %>% 
  ggplot(aes(x=N.Transports,y=N.Accidents)) + geom_point(alpha = 0.5) + stat_smooth(method="lm") + 
  geom_text_repel(aes(label=District.Name)) +
  labs(x='', y='', title = "Number of Accidents against Transports in the District") +
  theme(plot.title = element_text(size=14, margin=margin(b=14))) 

p1 + p2
grid::grid.draw(grid::textGrob("Number of Transports", y=0.05, x=0.52))

```
It would be naïve of to assume it would all be explained by only this one variable. The number of stations could very well be connected to the physical area of the district/neighborhood, which obviously means a higher probability of accidents.
However the correlations seem to be stronger than the population correlation the we showed above ($\rho = 0.70/0.74, p < 0.05$ vs $\rho= 0.76/0.83, p < 0.05$ for Neighborhood/District respectively), so it still seems like a good indicator.
```{r corr_trans_x_acc_tests, results='hide'}

print("Number of Transports in neighboorhood vs N Accident")
cor.test(full_neighborhood_dt$N.Transports, full_neighborhood_dt$N.Accidents, method="spearman", exact=FALSE)

print("Number of Transports in district vs N Accident")
cor.test(full_district_dt$N.Transports, full_district_dt$N.Accidents, method="spearman")
```

### Bus stops as confounding variable?

One could argue that the number of buses in an area is the true driving factor instead of other transport stops. The latter serve mostly off-road vehicles like trains, underground or boats, but they are probably connected to bus stops that do directly influence street-traffic. 

Before making conclusions however, it is necessary to investigate the different associations between transport or bus stops and accidents. 

```{r cof_var, message=FALSE, out.width="49%"}

ggplot(full_neighborhood_dt) + geom_point(aes(N.Bus.Stops, N.Accidents, color = "Bus")) + 
  geom_point(aes(N.Transports, N.Accidents, color = "Transports")) + 
  geom_smooth(aes(N.Bus.Stops, N.Accidents, color = "Bus"), method = "lm") + 
  geom_smooth(aes(N.Transports, N.Accidents, color = "Transports"), method = "lm") +
  labs(x = "Number of stops", y="Number of accidents", title = "Correlation between Stops and Accidents by neighborhood", color="Stop type") +
  guides(color=guide_legend(override.aes=list(fill=NA))) +
  theme(legend.position = c(0.93, 0.1)) 

ggplot(full_neighborhood_dt, aes(N.Transports, N.Bus.Stops)) + geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") + 
  labs(title = "Correlation between Transport and Bus stops by neighborhood", x="Number of transport stops", y="Number of bus stops")


## correlation coefficients
transport_acc_rho <- cor.test(full_neighborhood_dt$N.Transports, full_neighborhood_dt$N.Accidents, method = "spearman")
bus_acc_rho <- cor.test(full_neighborhood_dt$N.Bus.Stops, full_neighborhood_dt$N.Accidents, method = "spearman")
transport_bus_rho <- cor.test(full_neighborhood_dt$N.Transports, full_neighborhood_dt$N.Bus.Stops, method = "spearman")

```

The two graphs above show that transport is positively associated with accidents ($\rho=0.76, p<0.05$) as well as bus stops and accidents ($\rho = 0.66, p<0.05$). This is true also transports and bus stops ($\rho=0.52,p<0.05$). It is possible that bus stops are a confounding variable in the correlation between transport stops and accidents. Thus, it is necessary to check the correlation again while keeping the number of bus stops constant.

```{r message=FALSE, fig.height = 5, fig.width=7, out.height="40%", fig.align = "center"}

full_neighborhood_dt[, bus_strata := with(full_neighborhood_dt, cut(N.Bus.Stops, breaks = c(0, 30, 60, 90, 137),  
                                labels = c("0-30","31-60","61-90", ">90")))] %>%
  ggplot(aes(N.Transports, N.Accidents)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm") + 
  facet_wrap(~ bus_strata, scales = "free") + 
  labs(x="Number of transport stops", y ="Number of accidents", title="Correlation Accidents by Transports controlled for the Number of Bus Stops")
  
```

In the above plot the number of bus stops are binned into 4 range categories. The correlation between accidents and transport stations seemingly persists even when conditioning for bus stops. However, in the last category only few datapoints remain and the resulting correlation is visibly influenced by the outlier having about 60 transport stations. 

Below is the result for the spearman correlation test between :

- accidents and transport, controlling for bus stops  
- accidents and bus stops, controlling for transport stops. 

```{r echo=FALSE, fig.keep="all", message=FALSE, running=FALSE}
res1 <- pcor.test(full_neighborhood_dt$N.Accidents, full_neighborhood_dt$N.Transports, full_neighborhood_dt$N.Bus.Stops, method = "spearman")
res2 <- pcor.test(full_neighborhood_dt$N.Accidents, full_neighborhood_dt$N.Bus.Stops, full_neighborhood_dt$N.Transports, method = "spearman")

cor_dt <- data.table(test = c("accidents vs. transports",
                    "accidents vs. bus_stops"))
cor_dt <- cbind(cor_dt, rbind(res1, res2))
#cor_dt
knitr::kable(cor_dt)
```


A positive correlation between accidents and transport stops of about 0.65 still remains and is statistically significant, even when controlling for bus stops.


### Afflux throught the day

If our hypothesis is that the accidents are due to the afflux of people, we would expect their distribution throughout the day or the week not to be uniform.

```{r acc_hourly, fig.show="hold", out.width="50%"}
ggplot(acc_dt, aes(x=Hour)) + geom_bar() +labs(y="Number of accidents", title="Number of accidents by Hour")

acc_dt %>%
  mutate(Weekday=factor(Weekday, levels=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  ggplot(aes(x=Weekday)) + geom_bar() +
  labs(title="Number of accidents by Weekday") +
  theme(axis.title.y = element_blank())
```

As we can see, there are three peaks in the number of accidents: around 8-9 o'clock, 13-14 o'clock, and 18-19 o'clock. This leads us to believe that the traffic is expecially high at those times, which seems logical since these hours correspond to the _going-to-work-time_, _lunch-time_ and _coming-back-from-work-time_.
This theory is further reinforced by the fewer of accidents during the weekend, days that people tend to spend at home.

For further visualization let's show the accident density first in general, and then facetted by the part of the day (Morning: 6-13 o'clock, Afternoon: 14-21 o'clock, Night: 22-5 o'clock).


```{r acc_density, fig.height=2.8, fig.align = "center"}
acc_dt %>%
  mutate(Part.of.the.day=factor(Part.of.the.day, levels=c("Morning", "Afternoon", "Night"))) %>%
  ggplot(aes(Longitude, Latitude)) + 
  stat_density_2d(aes(fill = ..level..), geom = "polygon", binwidth=51) +
  labs(title = "Density of accidents in Barcelona by Part of the Day", fill="Number of Accidents") + 
  facet_wrap(~Part.of.the.day) +
  scale_fill_viridis() +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        legend.title = element_text(size=8),
        title = element_text(size=8),
        panel.grid = element_blank()) 

```
What is interesting here, is that the density focuses into one central point, namely the Eixample disctrict (calculated from the Long and Lat variables). We can clearly see that accidents in areas leading from/to (especially south and north) the center get more dense during the morning and afternoon. This further proves the above theory.


### Part of the Day Prediction - Logistic Regression 

To better understand the nature of the accidents, a prediction task was made using logistic regression. In this prediction, part of the day of the accident (which was categorized as morning and non-morning, which includes "afternoon" and "night") was guessed using the features "district name", "weekday", "month", "mild injuries", "serious injuries", "victims", "vehicles involved". Morning is annotated as 0, and "Non-morning" is annotated as 1.

```{r log_reg, message = FALSE, results="hide", fig.keep="all", out.width="49%"}

# Selecting Features
encoded_dt <- acc_dt[, c(2,5,6,7,8,9,10,11,12,13)] 

# Predicting: part of the days morning, not-morning (afternoon, night)
# Features: district name, weekday, month, mild injuries, serious injuries, victims, vehicles involved 

# Encoding categorical data columns.
encoded_dt$District.Name <- as.numeric(as.factor(encoded_dt$District.Name))
encoded_dt$Weekday <- factor(encoded_dt$Weekday, level = c("Friday", "Thursday", "Wednesday", "Saturday", "Tuesday", "Monday","Sunday"), labels = c(5,4,3,6,2,1,7))
encoded_dt$Month <- factor(encoded_dt$Month, level = c("October","September", "December", "July", "May", "June","January", "April", "March", "November", "February", "August" ),
                         labels = c(10,9,12,7,5,6,1,4,3,11,2,8))
encoded_dt$Part.of.the.day = factor(encoded_dt$Part.of.the.day, level = c( "Morning", "Afternoon", "Night"), labels = c(0,1,1))



# Shuffling the data
encoded_dt <- encoded_dt[sample(1:nrow(encoded_dt)), ] 


# split as train-test data
create_train_test <- function(data, size = 0.8, train = TRUE) {
  n_row = nrow(data)
  total_row = size * n_row
  train_sample <- 1: total_row
  if (train == TRUE) {
    return (data[train_sample, ])
  } else {
    return (data[-train_sample, ])
  }
}


data_train <- create_train_test(encoded_dt, 0.8, train = TRUE)
data_test <- create_train_test(encoded_dt, 0.8, train = FALSE)

# Building Multiple logistic regression model

multi_logistic_model <- glm(Part.of.the.day ~ District.Name + Weekday + Month + Day + Hour + Mild.injuries + Serious.injuries + Victims + Vehicles.involved, data = data_train, family = "binomial")

# Prediction & Classification
predict_reg <- predict(multi_logistic_model, data_test, type = "response")
predict_reg <- ifelse(predict_reg >0.5, 1, 0)

# Evaluating model accuracy using confusion matrix
confusion_matrix <- data.table(table(data_test$Part.of.the.day, predict_reg))
plot_conf_matrix <- confusion_matrix %>%
  mutate(goodbad := ifelse(confusion_matrix$predict_reg == confusion_matrix$V2, "good", "bad")) %>%
  group_by(V2) %>%
  mutate(prop = N/sum(N))

# Calculating accuracy
missing_class <- mean(predict_reg != data_test$Part.of.the.day)
print(paste('Model Accuracy (using mean() function) :', 1 - missing_class))

# Plot Conf Matrix
# fill alpha relative to sensitivity/specificity by proportional outcomes within reference groups 
ggplot(plot_conf_matrix, aes(x = V2, y = predict_reg, fill = goodbad, alpha = prop)) +
  geom_tile() +
  geom_text(aes(label = N), vjust = .5, alpha = 1) +
  scale_fill_manual(values = c(good = "green", bad = "red")) +
  scale_x_discrete(labels=c("0" = "Morning", "1" = "Non-Morning")) + 
  scale_y_discrete(labels=c("0" = "Morning", "1" = "Non-Morning")) + 
  labs(title=paste0('Confusion Matrix (Overall Accuracy:  ', round(1 - missing_class, 3), ')'), x="True Label", y="Predicted Label") +
  guides(fill="none", alpha="none") + 
  theme_bw()

# Calculate ROC Curve and AUC. Plot afterwards. 
rocobj <- roc(data_test$Part.of.the.day, predict_reg)
auc <- round(auc(data_test$Part.of.the.day, predict_reg),4)

ggroc(rocobj, colour = 'steelblue', size = 2) +
  ggtitle(paste0('ROC Curve ', '(AUC = ', auc, ')')) 

```


# Conclusions

TODO:

One crucial conclusion derived from the data is that where people move to is as important (or more) as to where they live. 
In increasingly expanding cities, people start to move to the outskirts (due to different factors like house pricing),  traveling daily to the center to work. Although we can't be sure this is the reason for the number of accidents in Eixample, we can notice looking at the map that this is indeed the most touristic and central district of Barcelona.

TODO: maybe image of the map

